Evaluation of classical correlation functions from 2/3D images on CPU and GPU architectures: introducing CorrelationFunctions.jl

### 2022-10-04 paper-correlations.tex
Add mathematical introduction

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\linewidth]{images/corr.png}
  \caption[]{WTF?}
  \label{fig:functions}
\end{figure}

@@

The aim of this paper is to establish a general open-source solution allowing to
compute all classical correlation functions. This solution should be well
documented with all computational algorithms explained, it should also leverage
on recent advances in programming high-intensity computations and utilize both
CPU and GPU power. Our answer in the form of \code{CorrelationFunctions.jl}
package allows to compute numerous CFs easily under any operating system. The
manuscript describing the package is organized as follows: in \cref{sec:scan} we
provide all algorithmic details for CFs computation in directions (scanning
method). \cref{sec:map} presents all algorithmic details for computing full CFs
maps (map method, full 2-point statistics for a given image). In
\cref{sec:verification} we verify the all computations based on known analytical
solutions. \cref{sec:efficiency} deals with computational efficiency of our
code, we compare computational times for 2D and 3D images of different sizes for
scanning and map methods as evaluated using either CPU or GPU. In
\cref{sec:examples} some application examples are provided, including CFs
evaluation for multi-phase materials. Finally, \cref{sec:summary} summarizes all
results and outlines possible future improvements.

\section{Overview of supported functions and modules.}
\label{sec:overview}
\verb+CorrelationFunctions.jl+ is capable to compute the following correlation
functions:
\begin{itemize}
\item Lineal-path function $L_2$.
\item Two-point function $S_2$.
\item Cluster function $C_2$.
\item Surface-surface function $F_{ss}$.
\item Surface-void function $F_{sv}$.
\item Pore size function $P$.
\item Chord length function $p$.
\item Phase cross-correlation function $\rho_{ij}$
\end{itemize}

@@

\section{The map approach.}
\label{sec:map}

In this approach we actively use fast Fourier transform (FFT)
to compute some correlation function ($S_2, F_{ss}, F_{sv}, C_2$)
in all directions.
The FFT algorithm let us compute auto and cross-correlation fast,
especially on GPU.

@@

\subsection{Surface-void function}

Computation of surface-void function consists of three steps 
\textcolor{red}{[link to surface paper]}:
\begin{enumerate}
  \item extracting surface $M = \nabla I(a)$
  \item extracting void phase $V = I_v(a)$
  \item computing $F_{sv} = cross\_correlation(M, V)$
\end{enumerate}

>>>>

\newtheorem{definition}{Definition}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\linewidth]{images/corr.png}
  \caption[]{A schematic depiction of a binary porous media (pores are shown in
    color) with examples of positive events for surface $F_{ss}$, $F_{sv}$ and
    two-point probability $S_2$ correlation functions. The zoomed in area
    represents the difference between the true ``continuous'' interface in
    between pore and solid phases with pixelized ``digital'' interface emerging
    due to limited resolution of digital images.}
  \label{fig:functions}
\end{figure}

@@

The aim of this paper is to establish a general open-source solution allowing to
compute all classical correlation functions. This solution should be well
documented with all computational algorithms explained, it should also leverage
on recent advances in programming high-intensity computations and utilize both
CPU and GPU power. Our answer in the form of \code{CorrelationFunctions.jl}
package allows to compute numerous CFs easily on any operating system. The
manuscript describing the package is organized as follows: in \cref{sec:math} we
introduce a reader to the most common and useful correlation functions,
\cref{sec:map} presents all algorithmic details for computing full CFs maps (map
method, full 2-point statistics for a given image). In \cref{sec:scan} we
provide all algorithmic details for CFs computation in directions (scanning
method). In \cref{sec:verification} we verify the all computations based on
known analytical solutions. \cref{sec:efficiency} deals with computational
efficiency of our code, we compare computational times for 2D and 3D images of
different sizes for scanning and map methods as evaluated using either CPU or
GPU. In \cref{sec:examples} some application examples are provided, including
CFs evaluation for multi-phase materials. Finally, \cref{sec:summary} summarizes
all results and outlines possible future improvements.

\section{Introduction to the correlation functions}
\label{sec:math}
In practice porous media are often described with special descriptors called
``correlation functions''. In this section we provide a brief introduction to
these functions which can also be found in \cite{Torquato_book}. The functions
supported by our package are:
\begin{itemize}
\item Lineal-path function $L_2$.
\item Two-point function $S_2$.
\item Cluster function $C_2$.
\item Surface-surface function $F_{ss}$.
\item Surface-void function $F_{sv}$.
\item Pore size function $P$.
\item Chord length function $p$.
\item Phase cross-correlation function $\rho_{ij}$
\end{itemize}

One of basic correlation functions is a lineal path function which is defined as
follows for homogenous media:
\begin{definition}
  Lineal path function $L_2^{(i)}(\mathbf{r})$ equals to probability that a
  vector $\mathbf{r}$ lies wholly in phase $i$ when randomly thrown into the
  sample.
\end{definition}
If a sample is isotropic a vector $\mathbf{r}$ can be replaced with its length
$r$ and the definition becomes:
\begin{definition}
  Lineal path function (for isotropic media) $L_2^{(i)}(r)$ equals to
  probability that a line segment of length $r$ lies wholly in phase $i$ when
  randomly thrown into the sample.
\end{definition}

Another function of big importance is a two-point correlation function $S_2$. A
definition for homogenous media is as follows:
\begin{definition}
  Two point function $S_2^{(i)}(\mathbf{r})$ equals to probability that both ends
  of a vector $\mathbf{r}$ lie in phase $i$ when the vector is randomly thrown
  into the sample.
\end{definition}
For isotopic media a vector $\mathbf{r}$ can be again reduced to a scalar value
$r$ (i.e. a correlation function depends only on length of a vector and not on
its direction.) Mathematically this definition can be expressed with the
following equation:
\begin{equation}
  S_2^{(i)}(\mathbf{r}) = \langle I^{(i)}(\mathbf{x}) I^{(i)}(\mathbf{x} +
  \mathbf{r}) \rangle
  \label{eq:s2-def}
\end{equation}
where $I^{(i)}$ is an indicator function for a set of all points belonging to
the phase $i$ and $\langle \cdots \rangle$ is an average over all xs.

A function closely related to $S_2$ is phase cross-correlation function
$\rho_{ij}$ which is defined with the following equation:
\begin{equation}
  \rho_{ij}(\mathbf{r}) = \langle I^{(i)}(\mathbf{x}) I^{(j)}(\mathbf{x} +
  \mathbf{r}) \rangle
  \label{eq:cross-def}
\end{equation}

The next function is a cluster function $C_2$.
\begin{definition}
  Cluster function $C_2^{(i)}(\mathbf{r})$ equals to probability that both
  ends of a vector $\mathbf{r}$ lie in the same cluster when the vector is
  randomly thrown into the sample. A cluster is a set of points of the same
  phase $i$ where any to points can be connected by a path lying entirely in
  that phase.
\end{definition}

The next two functions describe surface of a sample. These functions are
surface-surface and surface-void functions. For homogenous media their
definitions are:
\begin{definition}
  Surface-surface function $F_{ss}^{(i)}(\mathbf{r})$ equals to probability that
  both ends of a vector $\mathbf{r}$ touch the interface of phase $i$ when the
  vector is randomly thrown into the sample.
\end{definition}
and
\begin{definition}
  Surface-void function $F_{sv}^{(i)}(\mathbf{r})$ equals to probability that
  one end of a vector $\mathbf{r}$ touches the interface of phase $i$ and the
  other end touches the void phase when the vector is randomly thrown into the
  sample.
\end{definition}
If we introduce an interface indicator function
\begin{equation*}
  M^{(i)}(\mathbf{r}) = | \nabla I^{(i)}(\mathbf{r}) |
\end{equation*}
we can define these functions mathematically as
\begin{align}
  F_{ss}^{(i)}(\mathbf{r}) &= \langle M^{(i)}(\mathbf{x}) M^{(i)}(\mathbf{x} +
  \mathbf{r}) \rangle \label{eq:fss-def} \\ 
  F_{sv}^{(i)}(\mathbf{r}) &= \langle M^{(i)}(\mathbf{x}) I^{(void)}(\mathbf{x}
  + \mathbf{r}) \rangle \label{eq:fsv-def}
\end{align}
As always, there are versions of these functions applicable to scalar argument
for isotropic media.

Pore size function $P(\delta)$ is a probability density function defined as
follows:
\begin{definition}
  $P(\delta)d\delta$ = Probability that a randomly chosen point in a set of points
  belonging to the void phase lies at a distance between $\delta$ and $\delta + d\delta$
  from the nearest point on the pore-solid interface.
\end{definition}

Finally, a chord length function $p^{(i)}(r)$ is defined as follows:
\begin{definition}
$p^{(i)}(r)dz$ = Probability of finding a chord of length between $r$ and $r+dr$
in phase $i$. Chords are all of the line segments between intersections of an
infinitely long line with the two-phase interface.
\end{definition}

\section{Overview of supported functions and modules.}
\label{sec:overview}
\verb+CorrelationFunctions.jl+ is capable to compute the following correlation
functions:

@@

\section{The map approach.}
\label{sec:map}
In this approach we actively use fast Fourier transform (FFT)
to compute some correlation function ($S_2, F_{ss}, F_{sv}, C_2$)
in all directions.
The FFT algorithm let us compute auto and cross-correlation fast,
especially on GPU.

@@

\subsection{Surface-void function}
Computation of surface-void function consists of three steps 
\textcolor{red}{[link to surface paper]}:
\begin{enumerate}
  \item extracting surface $M = \nabla I(a)$
  \item extracting void phase $V = I_v(a)$
  \item computing $F_{sv} = cross\_correlation(M, V)$
\end{enumerate}
